---
title: "Worksheet 5a"
author: "Natalie Graham"
date: "2025-08_01"
output: pdf_document
---
## Introduction

In this exercise, we will work through the Peake example, which examines the relationship between area (dependent variable) and abundance of non-mussel invertebrates (response variable: INDIV) from 35 intertidal mussel clumps. Students will work in pairs from Steps 1-6 in Logan to gain experience with diagnostics, transformations, model interpretation, and new techniques for making a publishable scatterplot to summarize findings. We will work through Step 7 together as a class.

## Steps 1-5

First, read in the "peake" dataset and install the necessary packages.

```{r setup, include=FALSE}
#note you will need to run install.packages("gplots")
library(gplots)
library(car)
library(ggplot2)
```

Read in peake
```{r}
peake <- read_excel("~/data/peake.xlsx")

```

View the data and create scatterplots
```{r}
peake
scatterplot(INDIV ~ AREA, data = peake)
scatterplot(log10(INDIV) ~ log10(AREA), data = peake)
```

Fit a linear model and summarize it.
```{r}
peake.lm <- lm(log10(INDIV) ~ log10(AREA), data = peake)
summary(peake.lm)
```

Plot the model and check influence measures.
```{r}
plot(peake.lm)
influence.measures(peake.lm)
summary(peake.lm)
```

Create a scatterplot with custom axes and labels using base R
```{r}
plot(INDIV ~ AREA, peake, pch = 16, axes = FALSE, xlab = "", ylab = "", log = "xy")
axis(1, cex.axis = .8)
mtext(text = expression(paste("Mussel clump area", (mm^2))), side = 1, line = 3)
axis(2, las = 1)
mtext("Number of Individuals", side = 2, line = 3)
abline(peake.lm)
text(30000, 30, "log of indiv vs area")
text(1000, 30, "log of indiv vs area")

```

Try plotting the same plot with ggplot
```{r}


```

# Predicting Values
Use the fitted linear model to predict the number of individuals at given areas. Use the predict function to get the fitted values and confidence intervals. 

The predicted values from predict(peake.lm, newdata) will be on the log10 scale. To convert these predictions back to the original scale of INDIV, you need to exponentiate them:
original_scale_predictions <- 10^predict(peake.lm, newdata)

*predict*
The predict function in R is used to generate predicted values from a fitted model. When you use predict(peake.lm, newdata), it returns the predicted values on the same scale as the response variable used in the model.

Thus use *10^predict* or another transformation to predict...
When you are transforming the predicted values from the transformed scale back to the original scale. 

```{r}
# Predict the fitted values and confidence intervals
predicted_values <- 10^predict(peake.lm, data.frame(AREA = c(8000, 10000)))
predicted_values

```

Fit confidence intervals around these predictions.

```{r}
confidence_intervals <- 10^predict(peake.lm, data.frame(AREA = c(8000, 10000)), interval = "prediction")
confidence_intervals

```

Add 95% confidence bands to the plot.

```{r}
plot(log10(INDIV) ~ log10(AREA), data = peake, pch = 16)
x <- seq(min(peake$AREA), max(peake$AREA), length.out = 1000)
y <- predict(peake.lm, data.frame(AREA = x), interval = "confidence")
matlines(log10(x), y, lty = 1, col = 1)

```

Here’s how you can create the same plot using ggplot2 (so easy!):

```{r}

# Create the base plot
ggplot(peake, aes(x = log10(AREA), y = log10(INDIV))) +
  geom_point(shape = 16) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = expression(paste("log10(Mussel clump area ", (mm^2), ")")),
       y = "log10(Number of Individuals)") +
  theme_minimal()
```
So with ggplot...
geom_smooth(method = "lm", se = TRUE, color = "blue", fill = "lightblue") adds the linear regression line with 95% confidence bands. The fill parameter specifies the color of the confidence bands. But To change the confidence level to 99% in ggplot2, you can use the level parameter within geom_smooth (e.g. level = 99)

