---
title: "Homework1"
output:
  pdf_document: default
  html_document: default
---

# Case study 1. 

## Init

```{r}
# install.packages("car")
library(readxl)
beak.finch <- read_excel("/Users/kanoalindiwe/Downloads/Projects/playground/Datasets/beak.finch.xlsx")
year.cat <- as.factor(beak.finch$year) 
```
## 1. E
```{r}
summary(year.cat)
```

## 2. A/C
```{r}
tapply(beak.finch$depth, beak.finch$year, summary)["1976"]
```

## 3. B
```{r}
tapply(beak.finch$depth, beak.finch$year, summary)["1978"]
```

## 4. C
```{R}
hist(beak.finch$depth[beak.finch$year == 1976],
     main = "Beak depth in 1976",
     xlab = "Beak depth")

hist(beak.finch$depth[beak.finch$year == 1978],
     main = "Beak depth in 1978",
     xlab = "Beak depth",)
```

## 5. A (however, the whiskers go to 1.5 times the IQR)
```{R}
boxplot(beak.finch$depth ~ beak.finch$year)
```

## 6. A
```{r}
t.test(beak.finch$depth ~ beak.finch$year)
```

## 7. A

## 8. A

# Case study 2.

## Init

```{r}
macnally <- read_excel("/Users/kanoalindiwe/Downloads/Projects/playground/Datasets/macnally.xlsx")
```

## 1. EYR density is approximately normal, GST density is not normally distributed.
```{r}
boxplot(macnally$Density ~ macnally$Species, xlab="Species", ylab="Density")
hist(macnally$Density[macnally$Species=="GST"], xlab="Density", main="Density Histogram (GST)")
hist(macnally$Density[macnally$Species=="EYR"], xlab="Density", main="Density Histogram (EYR)")
```
## 2. Two sided test: failed to reject null that mean is different from 5.7. Less than test: Successfully rejected null. Greater than test: Failed to reject null.
```{r}
t.test(macnally$Density[macnally$Species=="GST"], mu=5.7)
t.test(macnally$Density[macnally$Species=="GST"], mu = 5.7, alternative = "less")
t.test(macnally$Density[macnally$Species=="GST"], mu = 5.7, alternative = "greater")
```

## 3. We fail to reject the null, there forfor we cannot say the data have significantly different means.
```{r}
eyr <- macnally$Density[macnally$Species == "EYR"]
gst <- macnally$Density[macnally$Species == "GST"]
t.test(eyr, gst, var.equal = TRUE)
t.test(eyr, gst, var.equal = FALSE)
```
## 4. There is not significant difference in medians.
```{r}
wilcox.test(eyr, gst, alternative = "two.sided")
```

# Case study 1 Questions:
1. Which function would you use to generate several basic summary stats for beak depth?
2. Basic summary stats of beak depth for the year 1976 are (hint, use tapply 
to get summary statistics by year):
3. Basic summary stats of beak depth for the year 1978 are:
4. Use the function "hist" to make histogram of finch beak depth for the years 1976 and 1978
(make separate histograms for each year). An appropriate description of these figures is:
5. Use the function "boxplot" to make a box plot of finch beak depth for the years 1976 and 1978
(show both years in a single figure containing two box plots). An appropriate description of this
figure is:
6. Based on your results from the previous steps, choose a suitable statistical test for comparing
beak size by year.
7. Use R to run a statistical test for comparing beak size by year. What is your statistical
conclusion?
8. How should you describe the biological conclusion from your statistical test? If you chose Wilcox above now run Welch t test

# Case study 2 Questions:
1. Examine assumptions of parametric tests by creating boxplots and histograms of both bird species. Be sure to add appropriate labels to each plot. Describe the data distribution.

2. Run a one sample t test for both a 2-sided and one-sided alternative (you specify the direction) comparing the mean for GST against a mu of 5.7. Report your results. Do you accept or reject the Null in each test?

3. Run a two sample t test comparing EYR to GST (first assume equal variances, then run another test assuming unequal variances). Report results for each. Do you accept or reject the Null?

4. Now run an appropriate non-parametric two-sample test
